Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2021-12-23 analysis/paper/paper.Rmd
added clarification of fish versus invertebrates

The largest uncertainty among model parameters was the proportion of the Northern Pikeminnow diet consisting of fishes (e.g., juvenile Chinook Salmon) versus other food items (e.g., invertebrates). Therefore, we varied the proportion of the diet consisting of fish versus invertebrates to range from 10-90%, in 10% intervals, for both categories. Energy densities for invertebrates was fixed at 3,000 J/g. Separate models were also run for a 77-day fall period from September 15 - November 30 and a 92-day period in the spring from March 1 - May 31 to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG. This resulted in 18 separate model runs (9 diet scenarios and 2 seasons). Each model run provided an estimate of the amount (grams) of fish consumed by an individual Northern Pikeminnow during the fall or spring time periods.

@@

To estimate the total biomass of DSR and NRR emigrants potentially consumed by all Northern Pikeminnow in Deadwater Slough, we then multiplied the grams of fish consumed by an individual Northern Pikeminnow  during each period (fall and spring) by the estimated population size for that period. The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed each period by dividing by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The weight of an average DSR and NRR emigrants was taken from fish caught at seven rotary screw traps located throughout the Upper Salmon MPG upstream of Deadwater Slough during the fall and spring time periods.

@@

We caught `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "n_fish"])`, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "n_fish"])`, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "n_fish"])` Northern Pikeminnow during the fall 2019, fall 2020, and spring 2021 efforts, respectively, for a total of `r prettyNum(sum(total_cpue$n_fish))` fish caught using hook-and-line angling. The CPUE, measured as number of Northern Pikeminnow caught per angler hour, was `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "catch_per_hr"], digits = 3)` for fall 2019, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "catch_per_hr"], digits = 3)` for fall 2020, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "catch_per_hr"], digits = 2)` for spring 2021 (Figure \@ref(fig:cpue-fig)). During the mark-recapture efforts, there were `r as.character(english(sum(N_mods[[3]][[1]]$m)))` recaptures in fall 2019 which included `r as.character(english(LP_mod[[3]][[1]]$m))` unique individuals (i.e., `r as.character(english(sum(N_mods[[3]][[1]]$m) - LP_mod[[3]][[1]]$m))` individual was captured more than once). There were an additional `r as.character(english(sum(N_mods[[3]][[3]]$m)))` recaptures in fall 2020 which included `r as.character(english(LP_mod[[3]][[3]]$m))` unique individuals. Summaries of parameters for both the single census and multiple census mark-recapture estimators are provided in Table \@ref(tab:abund-data-lp) and Table \@ref(tab:abund-data-schn), respectively.

@@

Given a mean population size of `r prettyNum(unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` from fall estimates and an estimate of `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")` for spring 2021, and assuming the median value from diet scenarios (i.e, 50% of their diet consists of fish), we'd estimate that `r prettyNum(round(median(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")` (95% CI: `r prettyNum(round(min(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")` - `r prettyNum(round(max(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")`) DSR and `r prettyNum(round(median(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")` (95% CI: `r prettyNum(round(min(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")` - `r prettyNum(round(max(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")`) NRR juveniles may be consumed in Deadwater Slough during peak emigrations. Table \@ref(tab:adult-imp) shows the estimated number of juvenile Chinook Salmon consumed by and individual Northern Pikeminnow and by the Deadwater Slough population across all diet scenarios and for both the fall (DSR) and spring (NRR) peak emigration seasons.

Across both seasons and all diet scenarios, we estimated the total juvenile outmigrants consumed to be between `r prettyNum(round(min(total_impacts$smolt_N), 0), big.mark = ",")` and `r prettyNum(round(max(total_impacts$smolt_N), 0), big.mark = ",")`. The estimated number of Chinook Salmon consumed by a Northern Pikeminnow begins to taper off betwen the 30-90% diet scenarios. So although we are uncertain of the diet proportions, the impact of that uncertainty is somewhat constrained.

>>>>

The largest uncertainty among model parameters was the proportion of the Northern Pikeminnow diet consisting of fish (e.g., juvenile Chinook Salmon) versus other food items (e.g., invertebrates). Therefore, we varied the proportion of the diet consisting of fish versus invertebrates to range from 10-90%, in 10% intervals, for both categories. Energy densities for invertebrates was fixed at 3,000 J/g. We estimate the g of (any) fish consumed for each model run. For simplicity, we assume that any fish prey have the same energy densities as juvenile Chinook Salmon and we do not differentiate juvenile Chinook Salmon from other fish prey than may be present in Deadwater slough (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner). Separate models were also run for a 77-day fall period from September 15 - November 30 and a 92-day period in the spring from March 1 - May 31 to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG. This resulted in 18 separate model runs (9 diet scenarios and 2 seasons). Each model run provided an estimate of the amount (grams) of fish consumed by an individual Northern Pikeminnow during the fall or spring time periods.

@@

To estimate the total biomass of fish potentially consumed by all Northern Pikeminnow in Deadwater Slough, we then multiplied the grams of fish consumed by an individual Northern Pikeminnow  during each period (fall and spring) by the estimated predator population size for that period. The total biomass was further converted to an estimate of the total number of fish consumed each period by dividing by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The weight of an average DSR and NRR emigrants was taken from fish caught at seven rotary screw traps located throughout the Upper Salmon MPG upstream of Deadwater Slough during the fall and spring time periods. This calculation assumes that all fish prey items consumed by Northern Pikeminnow are, on average, the same size as DSR and NRR emigrants.

@@

We caught `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "n_fish"])`, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "n_fish"])`, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "n_fish"])` Northern Pikeminnow during the fall 2019, fall 2020, and spring 2021 efforts, respectively, for a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` fish caught using hook-and-line angling. The CPUE, measured as number of Northern Pikeminnow caught per angler hour, was `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "catch_per_hr"], digits = 3)` for fall 2019, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "catch_per_hr"], digits = 3)` for fall 2020, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "catch_per_hr"], digits = 2)` for spring 2021 (Figure \@ref(fig:cpue-fig)). During the mark-recapture efforts, there were `r as.character(english(sum(N_mods[[3]][[1]]$m)))` recaptures in fall 2019 which included `r as.character(english(LP_mod[[3]][[1]]$m))` unique individuals (i.e., `r as.character(english(sum(N_mods[[3]][[1]]$m) - LP_mod[[3]][[1]]$m))` individual was captured more than once). There were an additional `r as.character(english(sum(N_mods[[3]][[3]]$m)))` recaptures in fall 2020 which included `r as.character(english(LP_mod[[3]][[3]]$m))` unique individuals. Summaries of parameters for both the single census and multiple census mark-recapture estimators are provided in Table \@ref(tab:abund-data-lp) and Table \@ref(tab:abund-data-schn), respectively.

@@

Given a mean population size of `r prettyNum(unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` from fall estimates and an estimate of `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")` for spring 2021, and assuming the median value from diet scenarios (i.e, 50% of their diet consists of fish), we'd estimate that Northern Pikeminnow can consume `r prettyNum(round(median(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")` (95% CI: `r prettyNum(round(min(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")` - `r prettyNum(round(max(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")`) fish during the DSR emigration and `r prettyNum(round(median(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")` (95% CI: `r prettyNum(round(min(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")` - `r prettyNum(round(max(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")`) fish during the NRR emigration in Deadwater Slough. Table \@ref(tab:adult-imp) shows the estimated number of fish consumed by an individual Northern Pikeminnow and by the population across all diet scenarios during both the fall (DSR) and spring (NRR) peak emigration seasons.

Across both seasons and all diet scenarios, we estimated the total number of fish consumed to be between `r prettyNum(round(min(total_impacts$smolt_N), 0), big.mark = ",")` and `r prettyNum(round(max(total_impacts$smolt_N), 0), big.mark = ",")` (Table \@ref(tab:adult-imp)). The estimated number of fish consumed by a Northern Pikeminnow begins to taper off between the 30-90% diet scenarios. So although we are uncertain of the diet proportions consisting of fish versus invertebrates, the impact of that uncertainty is somewhat constrained.
